Causal inference and co-evolutionary contingencies using dynamic phylogenetic models

### 2022-07-01 manuscript.qmd

Trait co-evolution simply describes *how* traits change over time. Such descriptions have value for understanding natural history, ancestral state reconstruction (using the present to retrodict the past), and can stimulate new research questions. However, simply inferring that traits changed together in the past does not tell us *why* they changed, or how the traits would respond to intervention, i.e., causality. Nonetheless, researchers in ecology and evolution are often motivated to study trait co-evolution because they believe these patterns reflect causal relations. We briefly review some popular classes of PCM and their limitations with respect to causal inference.

@@

#### When can be interpret the evolutionary regression as a causal effect?

@@

```
To understand when the evolutionary regression coefficient can be interpreted as a causal effect, we must first clarify what "causality" means. Following Pearl (some citations), we say that $X$ is a cause of $Y$ when $Pr(Y|do(X = x)) \neq Pr(Y|do(X = \textrm{not x}))$. The do() operator represents an *intervention* in which the value of some variable is fixed (for example, through assignment in a randomized experiment). If the probability distribution of $Y$ changes based on the value of $X$ that is assigned, then $X$ has a causal effect on $Y$. Note here that $X$ refers to a random variable while $x$ refers to some unit value of that variable.

@@

$$y_{ij} = \beta_0 + \beta_xx_{ij} + \epsilon_{ij}$$
$$ \epsilon \sim  \mathcal{N}(0, \boldsymbol{\Sigma})$$
$$\boldsymbol{\Sigma} = \sigma^2\boldsymbol{V} $$
Where $\mathcal{N}$ denotes a normal or Gaussian probability density function and $\sigma^2$ is the variance of $Y$ conditional on $X$. $\beta_x$ is an estimator of the evolutionary regression coefficient that takes into account the phylogenetic non-independence between tip species. The covariance between related species is proportional to the off-diagonals of the phylogenetic variance-covariance matrix $\boldsymbol{V}$. To construct $\boldsymbol{V}$, we need (1) a matrix of patristic distances between species (i.e., time since divergence from last common ancestor) and (2) an assumed model of trait evolution.

@@

$$ W_{T= t} - W_{T = 0} \sim\mathcal{N}(0,\sigma t) $$
Therefore, the variance for a given species or population $j$ is proportional to total branch length of the tree leading from root to tip, and the covariance between $j$ and any other species or population $k$ is proportional to their amount of shared history. For example, imagine a phylogeny with only $J$ = 3 tip species. The last common ancestor of all species is dated to T = 10,000 years ago, and the last common ancestor of 2 and 3 is dated to 0.5T = 5,000 years ago. The implied phylogenetic variance-covariance matrix ($\boldsymbol{V}$) under a Brownian motion model is thus:

@@

This scaling is often referred to a≠s a "branch length transformation," because shrinking the off-diagonals (shared evolutionary history) while holding the diagonals constant (independent evolution) is equivalent to "stretching" the terminal branches leading to the tips and "shrinking" the past co-evolutionary branches. $\lambda = 1$ indicates that the phylogenetic strength is as strong as expected under a Brownian motion model, while lower values means that a lower proportion of the trait variance can be attributed to shared evolutionary history. At the extreme of $\lambda = 0$, $\boldsymbol{V}_s$ becomes an identity matrix and PGLS reduces to standard least squares regression.

Brownian motion is the simplest and most widely used model, and often interpreted as "drift" across the phylogenetic tree. Other popular choices include Ornstein-Uhlenbeck, where change due is due to both random drift and "stabilizing selection" towards some optimal trait value, and Early Burst models of adaptive radiation where the rate of change ($\sigma$) decreases over time. In general, $\boldsymbol{V}$ is constructed by inputting a matrix of phylogenetic distances to some covariance function $C(x,x')$ such that $V_{[j,k]} = C(j,k)$.

Although PGLS is widely used in comparative biology, it has key limitations such as: assumption of Gaussian errors, prone to overfitting due to lack of parameter regularization, and cannot accommodate many common data structures including repeated measures of species trait values, missing data, and measurement error. In the next section we review more flexible regression approaches, as well as an important alternative method of estimating evolutionary associations.

@@

Phylogenetic generalized multilevel models (PGLMMs) extend the study of trait coevolution, leveraging analytic strategies developed in quantitative genetics to accommodate more complex, multivariate model designs with non-Gaussian outcomes [@hadfield2010general; @ives2010phylogenetic; @ives2011generalized]. Rather than explicitly scaling the off-diagonals of $\boldsymbol{V}_s$ with some parameter $\lambda$, phylogenetic signal is quantified as the ratio of the variance of phylogenetic random effects $\nu$ (also called 'mixed' and 'varying' effects) to total trait variance. For example, an observation of a Gaussian trait $Y$ from a species or population $j$ and predictor $X$ is modeled as:

@@

Where $\boldsymbol{I}$ is an identity matrix and thus the residuals $\epsilon$ are assumed conditionally independent. The "phylogenetic signal" is now estimated as $\lambda_{\textrm{PGLMM}} = \frac{\sigma_{\nu}^2}{\sigma_{\nu}^2 + \sigma_{\epsilon}^2}$ (adjusted for $X$) and equivalent to $\lambda_{\textrm{PGLS}}$, but with the greater flexibility afforded by PGLMMs. $\lambda_{\textrm{PGLMM}}$ is also sometimes referred to as "phylogenetic heritability" [@hadfield2010general].

>>>>

Trait co-evolution simply describes *how* traits change over time. Such descriptions have value for understanding natural history, ancestral state reconstruction (using the present to retrodict the past), and can stimulate new research questions. However, simply inferring that traits changed together in the past does not tell us *why* they changed, or how the traits would respond to intervention, i.e., causality. Nonetheless, researchers in ecology and evolution are often motivated to study trait co-evolution because they believe these patterns reflect causal relations. We briefly review some popular classes of PCM and their limitations with respect to causal inference, before introducing a new method which overcomes some of these limitations.

@@

#### When can we interpret the evolutionary regression as a causal effect?

@@

```

To understand when the evolutionary regression coefficient can be interpreted as a causal effect, we must first clarify what "causality" means. Following Pearl (some citations), we say that $X$ is a cause of $Y$ when $Pr(Y|do(X = x)) \neq Pr(Y|do(X = \textrm{not x}))$. The do() operator represents an *intervention* in which the value of some variable is fixed (for example, through assignment in a randomized experiment). If the probability distribution of $Y$ changes based on the value of $X$ that is assigned, then $X$ has a causal effect on $Y$. Note here that $X$ refers to some variable while $x$ refers to some unit value of that variable.

@@

$$y_{ij} = \beta_0 + \beta_xx_{ij} + \epsilon_{ij}$$ $$ \epsilon \sim  \mathcal{N}(0, \boldsymbol{\Sigma})$$ $$\boldsymbol{\Sigma} = \sigma^2\boldsymbol{V} $$ Where $\mathcal{N}$ denotes a normal or Gaussian probability density function and $\sigma^2$ is the variance of $Y$ conditional on $X$. $\beta_x$ is an estimator of the evolutionary regression coefficient that takes into account the phylogenetic non-independence between tip species. The covariance between related species is proportional to the off-diagonals of the phylogenetic variance-covariance matrix <!--# isn't this also often called A? doesn't matter i guess.. --> $\boldsymbol{V}$. To construct $\boldsymbol{V}$, we need (1) a matrix of patristic distances between species (i.e., time since divergence from last common ancestor) and (2) an assumed model of trait evolution.

@@

$$ W_{T= t} - W_{T = 0} \sim\mathcal{N}(0,\sigma t) $$ Therefore, the variance for a given species or population $j$ is proportional to the total branch length of the tree leading from root to tip, and the covariance between $j$ and any other species or population $k$ is proportional to their amount of shared history. For example, imagine a phylogeny with only $J$ = 3 tip species. The last common ancestor of all species is dated to T = 10,000 years ago <!--# maybe make it 1 million? more realistic for species -->, and the last common ancestor of 2 and 3 is dated to 0.5T = 5,000 years ago. The implied phylogenetic variance-covariance matrix ($\boldsymbol{V}$) under a Brownian motion model is thus:

@@

This scaling is often referred to as a "branch length transformation," because shrinking the off-diagonals (shared evolutionary history) while holding the diagonals constant (independent evolution) is equivalent to "stretching" the terminal branches leading to the tips and "shrinking" the past co-evolutionary branches. $\lambda = 1$ indicates that the phylogenetic strength is as strong as expected under a Brownian motion model, while lower values means that a lower proportion of the trait variance can be attributed to shared evolutionary history. At the extreme of $\lambda = 0$, $\boldsymbol{V}_s$ becomes an identity matrix and PGLS reduces to standard least squares regression.

Brownian motion is the simplest and most widely used model, and often interpreted as "drift" across the phylogenetic tree. Other popular choices include Ornstein-Uhlenbeck, where change is due to both random drift and "stabilizing selection" around some optimal trait value, as well as Early Burst models of adaptive radiation where the rate of change ($\sigma$) decreases over time. In general, $\boldsymbol{V}$ is constructed by inputting a matrix of phylogenetic distances to some covariance function $C(x,x')$ such that $V_{[j,k]} = C(j,k)$.

Although PGLS is widely used in comparative biology, it has several key limitations. These include the assumption of Gaussian errors, proneness to overfitting due to a lack of parameter regularization, and failure to accommodate many common data structures such as repeated measures of species trait values, missing data, and measurement error. In the next section we review more flexible regression approaches, as well as an important alternative method of estimating evolutionary associations.

@@

Phylogenetic generalized multilevel (or mixed-effects) models (PGLMMs) extend the study of trait coevolution by leveraging analytic strategies developed in quantitative genetics to accommodate more complex, multivariate model designs with non-Gaussian outcomes [@hadfield2010general; @ives2010phylogenetic; @ives2011generalized] <!--# there's also Lynch 1991 METHODS FOR THE ANALYSIS OF COMPARATIVE DATA IN EVOLUTIONARY BIOLOGY, Evolution -->. Rather than explicitly scaling the off-diagonals of $\boldsymbol{V}_s$ with some parameter $\lambda$, phylogenetic signal is quantified as the ratio of the variance of phylogenetic random effects $\nu$ (also called 'mixed' or 'varying' effects) to total trait variance. For example, an observation of a Gaussian trait $Y$ from a species or population $j$ and predictor $X$ is modeled as:

@@

Where $\boldsymbol{I}$ is an identity matrix and thus the residuals $\epsilon$ are assumed conditionally independent. The "phylogenetic signal" is now estimated as $\lambda_{\textrm{PGLMM}} = \frac{\sigma_{\nu}^2}{\sigma_{\nu}^2 + \sigma_{\epsilon}^2}$ (adjusted for $X$) and equivalent to $\lambda_{\textrm{PGLS}}$, but with the greater flexibility afforded by PGLMMs. $\lambda_{\textrm{PGLMM}}$ is also sometimes referred to as "phylogenetic heritability" [@hadfield2010general], since it is calculated in the same way as heritability based on pedigrees in quantitative genetics.
